Topological Equivariant Artist Model

### 2021-03-31 paper/paper.tex
still need to flesh out intro

%% the only exception to this rule is the \firstsection command
\firstsection{Introduction}
\maketitle
The underlying structure and semantics of visualization are by definition reflected in visual representations \cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity to external concepts \cite{byrneAcquiredCodesMeaning2016}. Visualization tools allow practioners to automate these transformations, but most visualization tools are either explicitly tuned to specific continuity \cite{HeerSoftware2006} or define every interface relative to its expected data structure. The former approach is often too restrictive for a general purpose tool, while the latter often results in a library that can feel inconsistent or disjointed. Motivated by the challenge of rearchitecturing a general purpose visualization library, we developed the Topological Equivariant Artist Model (TEAM) to express the constraints of preserving data continuity and translating data properties into a visual representation. The contribution of TEAM is 
\begin{enumerate}
  \item topology preserving relationship between data and graphic via continuous maps
  \item property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
  \item functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model.
\end{enumerate}

\section{Related Work}
The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, and the notion that the properties of the data and visual representation match is the basis of most evaluations of visualization. Expressiveness, as defined by Mackinlay \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987, mackinlayAutomatingDesignGraphical1986} is a measure of how much of the structure any map can encode. A fully expressive component is one that is equivariant since it has preserved all the structure in the data. Models of visualization evaluate this equivariance and how elements built in the model can be composed to build more complex visualizations. Mackinlay's \textit{A Presentation Tool} (APT) introduced the notion of visualizations having syntax and semantics \cite{mackinlayAutomatingDesignGraphical1986} and Wilkenson described the grammar of this language \cite{wilkinsonGrammarGraphics2005}. This grammar oriented approach allows users to describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building those elements. This same limitation is explicitly stated in the functional dependency model of visualization developed by Sugibuchi\cite{sugibuchiFramwork2009} and evident in Vickers' category theory oriented framework in which semiotics are commutative. The algebraic process model by Kindlmann and Scheideggar proposes that the data and visualization transformations are commutative; while similar to our framework, it is missing any explicit mention of continuity.

Much of visualization algorithm design is oriented around continuity, as described in Tory and Möller's taxonomy\cite{toryRethinkingVisualizationHighlevel2004}. For example, the relational database is core to tools influenced by APT, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, Vega\cite{satyanarayanDeclarativeInteractionDesign2014} and Altair\cite{vanderplasAltairInteractiveStatistical2018}. Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and the digital humanities oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro; the need to visualize and manipulate graphs has spawned tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008}. Neither the table nor image nor graph model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} and D3 \cite{bostockDataDrivenDocuments2011} explicitly carry around different data representations for all the different types of visualizations they support.  Where libraries with a single core data structure have very consistent APIs, VTK, D3 and Matplotlib APIs can be rather inconsistent as every visualization has a different notion of how the data is structured.

Fiber bundle and sheafs are generalized abstractions proposed by Butler to encode the continuity of the data separately from the properties of each variable  \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}. Since Butler's model lacks a robust way of describing variables, we fold in Spivak's Simplicial formulation of databases \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES} to incorporate a schema like description of the variables. One way of describing the binding between the scheme and the continuinity is using the notion of structural \textit{keys} with associated \textit{values} proposed by Munzner\cite{munznerVisualizationAnalysisDesign2014}. Unlike Munzner's model where the semantic meaning of the key is tightly coupled to the index of the value, our model considers keys to be a reference to topology. This allows the metadata to be altered, for example by changing the coordinate system or time resolution, without imposing new semantics on the underlying structure.

\section{Topological Equivariant Artist Model}
\label{sec:math}
We introduce the notion of an artist $\mathscr{\vartist}$ as an equivariant map
\begin{equation}
    \label{eq:artist}
    \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
\end{equation}
from data $\mathscr{\dtotal}$ to graphic $\mathscr{\gtotal}$. To formalize the structure the artist is preserving, we describe how we model data (\autoref{sec:math:data}) and graphics (\autoref{sec:math:graphic}) as fiber bundles. We then discuss the equivariant maps that make up $\mathscr{\vartist}$, which are a map from  graphic continuity to data continuity (\autoref{sec:math:graphic:base}), data components to visual components (\autoref{sec:math:artist:nu}) and visual components to graphic (\autoref{sec:math:artist:q}).

@@

\subsubsection{Fiber Space: Variables}
\label{sec:math:data:fiber}
To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES} that binds the components of the fiber to variable names. Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_m\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset. He then defines the single variable set \fttype\ 
\begin{equation}
    \label{eq:data_types}
    \begin{tikzcd}
        \fttype \arrow[r] \arrow[d, "\pi_{\fsection}"'] & \ftotal \arrow[d, "\pi"] \\
        \fnames \arrow[r, "\fsection"']                          & \ftypes       
    \end{tikzcd}
\end{equation}
which is \ftotal\ restricted to objects of type \ftype\ bound to variable name \fname. The \fttype\ lookup is by name to specify that every component is distinct, since multiple components can have the same type \ftype. Given \fsection, the fiber for a one variable dataset is
\begin{equation}
    \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
\end{equation}
where \fsection\ is the schema binding variable name \fname\ to its datatype \ftype. A dataset with multiple variables has a fiber that is the cartesian cross product of $\ftotal_{\fsection}$ applied to all the columns:
\begin{equation}
F = \ftotal_{\fsection(\fname_{1})}\times \ldots \ftotal_{\fsection(\fname_{i})} \ldots\times \ftotal_{\fsection(\fname_{n})}
\end{equation}
which is equivalent to 
\begin{equation}
    \label{eq:math:data:fiber:decompose}
    \dfiber= \dfiber_{0} \times \ldots \times \dfiber_{i}\times\ldots\times \dfiber_{n}
\end{equation}
which allows us to decouple \dfiber\ into components $\dfiber_i$. Each component of \dfiber\ is a dimension of the topological fiber space and is specified by a tuple of the form (\fname, \ftype, \(\ftotal_{\fsection(\fname)}\)). 
\begin{figure}[htb]
  \centering 
  \includegraphics[width=\columnwidth]{fiber.png}
  \caption{These two datasets have the same base space \dbase. The plane is a representation of the fiber $\dfiber=\reals\times\reals$ for the variables (time, temperature),  while the cube is the fiber $\reals\times\realsp\times\reals$ associated with (time, wind=(speed, direction))}
  \label{fig:math:data:fiber}
\end{figure}
In \autoref{fig:math:data:fiber} the plane fiber has components (\emph{time}, \texttt{datetime}, \reals) and (\emph{temperature}, \texttt{float}, \reals), while the cube fiber has components (\emph{time}, \texttt{datetime}, \reals) and (\emph{wind}, \texttt{wind}, \(\realsp\times\reals)\)) which encodes (\emph{speed}, \emph{direction}).

@@

A monoid \cite{Monoid2021} $\monoid$ is a set with an associative binary operator $\ast:\monoid \times \monoid\rightarrow \monoid$. A monoid has an identity element $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$. As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,nlab:action} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
\begin{align*}
    \textbf{associativity}\;& \text{for all } f,g \in \monoid_i \text{ and } x\in \dfiber_i,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
    \textbf{identity}\;& \text{for all } x\in \dfiber_i, e\in \monoid_i,\,  e\bullet x = x 
\end{align*}
As with the fiber \dfiber\, the total monoid space \monoid\ is the cartesian product
\begin{equation}\label{eq:math:data:monoid:decompose}
\monoid= \monoid_{0} \times \ldots \times \monoid_{i}\times \ldots \times\ldots \monoid_{n}
\end{equation}
of each monoid $\monoid_{i}$ on $\dfiber_{i}$.  The monoid is also added to the specification of the fiber $(\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)$

Defining the monoid actions on the components serves as the basis for identifying the invariance\cite{kindlmannAlgebraicProcessVisualization2014} that must be preserved in the visual representation of the component. A secondary advantage of defining structure in terms of monoids is that they are commonly found in functional programming because they specify compositions of transformations \cite{yorgeyMonoidsThemeVariations, stievenMonadJustMonoid2020}.

\subsubsection{Base Space: Continuity}
\label{sec:math:data:base}
 The base space \dbase\ acts as an indexing space, as emphasized by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, to express how the records in \dtotal\ are connected to each other and does not itself store any of the components of the data. For example, \dbase\ can encode that the timeseries data described in \autoref{sec:math:data:fiber} is continuous, but is doing so independent of the timestamps. Every $\dbasepoint \in \dbase$ has a corresponding fiber $\dfiber_k$ because \dbase\ is the quotient space\cite{QuotientSpaceTopology2020,aurouxMath131Introduction} of \dtotal.

 \begin{figure}[htb]
  \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
  \includegraphics[width=\columnwidth]{base.png}
  \caption{These two datasets have the same fiber space  $\dfiber=\reals\times\reals$, but the left dataset has a discrete continuity while the right is 1D continuous over the interval $\left[0,1\right]$.}
  \label{fig:math:data:base}
 \end{figure}

 As illustrated in ~\autoref{fig:math:data:base}, \dbase\ can have any number of dimensions, can be continuous or discrete, and is somewhat independent of the dimensions of the fiber. As with \autoref{eq:math:data:fiber:decompose} and \autoref{eq:math:data:monoid:decompose},  we can decompose the total space into component bundles
 \begin{equation}\label{eq:math:data:base:decompose}
     \pi:\dtotal_1\oplus\ldots\oplus \dtotal_i \oplus\ldots \oplus \dtotal_n \rightarrow \dbase
 \end{equation}
 such that \(\monoid_{i}\) acts on component bundle \(\dtotal_i\). The \dbase\ remains the same because the connectivity of records does not change just because there are fewer components in each record. By encoding this continuity in the model as \dbase\, the data model now explicitly carries information about its structure such that the implicit assumptions of the visualization algorithms are now explicit. The explicit topology is a concise way of distinguishing visualizations that appear identical, for example heatmaps and images.

 \subsubsection{Section: Values}
 \label{sec:math:data:section}
 \begin{figure}[htb]
  \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
  \includegraphics[width=\columnwidth]{fiberbundle.png}
  \caption{Each section in the fiber bundle is a unique continuous map from base space to fiber encoding the set of records in the dataset. The two sections, $\dsection^{(1)}$ and $\dsection^{(2)}$, shown here are in the set of global sections $\Gamma(\dtotal)$.}
  \label{fig:math:data:section}
 \end{figure}

 While the projection function $\pi:\dtotal \rightarrow\dbase$ ties together the base space \dbase\ with the fiber \dfiber, a section $\dsection: \dbase\rightarrow \dtotal$ encodes a dataset. A section function takes as input location $\dbasepoint \in \dbase$ and returns a record $\delement \in \dtotal$. For any fiber bundle, there exists a map
 \begin{equation}
     \begin{tikzcd}
         \dfiber \arrow[r, hook] & \dtotal \arrow[d, "\pi"'] \\
                           & \dbase \arrow[u, "\dsection"', bend right]
     \end{tikzcd}
 \end{equation}
  such that $\pi(\dsection(\dbasepoint)) = \dbasepoint$. The set of all global sections is denoted as $\Gamma(\dtotal)$. As illustrated in \autoref{fig:math:data:section}, the section is a continuous mapping from a location $\dbasepoint \in \dbase$ on the base space to a record $\delement \in \dfiber$ in the fiber. Assuming a trivial fiber bundle $\dtotal = \dbase \times \dfiber$, the section is 
 \begin{equation}
     \label{eq:section_return}
     \dsection(\dbasepoint) = (\dbasepoint, (g_{\dfiber_{0}}(\dbasepoint), \ldots, g_{\dfiber_{n}}(\dbasepoint)))
 \end{equation}
 where $g: \dbase \rightarrow \dfiber$ is the index function into the fiber component. This formulation of the section also holds on locally trivial sections of a non-trivial fiber bundle. As with \autoref{eq:math:data:fiber:decompose} and \autoref{eq:math:data:base:decompose}, \dsection\ can be decomposed into components 
 \begin{equation}
 \label{eq:math:data:section:decompose}
 \dsection= (\dsection_0,\ldots, \dsection_i, \dots, \dsection_n) 
 \end{equation}
 where each section $\dsection_i$ maps into a record on a component $\dfiber_i \in \dfiber$. This allows for accessing the data component wise in addition to accessing the data in terms of its location over \dbase.

 \subsubsection{Sheafs}
 \label{sec:math:data:sheaf}
A sheaf is a mathematical structure for defining collections of objects\cite{ghristElementaryAppliedTopology2014, ghristHomologicalAlgebraData2018,urbanikBriefIntroductionSchemes} on mathematical spaces. On the fiber bundle \dtotal, we can describe a sheaf as the collection of local sections $\iota^{*}\dsection$
 \begin{equation}
     \label{eq:sheaf}
     \begin{tikzcd}
         \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             & \dtotal \arrow[d, "\pi"']                  \\
         U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] & \dbase \arrow[u, "\dsection"', bend right]
     \end{tikzcd}
 \end{equation}
which are sections of \dtotal\ pulled back over local neighborhood $U \subset \dtotal$ via the inclusion map \(\iota:\dtotal\rightarrow U\). The bookkeeping enabled by sheafs is necessary for navigation techniques such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006} and dynamically updated visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995}

@@

Just as the \dbase\ encodes the connectivity of the records in the data, we propose an equivalent \gbase\ that encodes the connectivity of the rendered elements of the graphic. Formally, we require that \dbase\ be a deformation retract\cite{RetractionTopology2020} of \gbase\ so that \dbase\ and \gbase\ have the same homotopy. The surjective map $\vindex: \gbase \rightarrow \dbase$ 
\begin{equation}
    \begin{tikzcd}
        \dtotal \arrow[d, "\pi"'] & \gtotal \arrow[d, "\pi"'] \\
        \dbase                   & \gbase \arrow[l, "\vindex"']
    \end{tikzcd}
\end{equation}
goes from region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\gbasepoint$. While \gbase\ must have the same continuity as \dbase\, it is sometimes the  thickened version shown in \autoref{fig:math:graphic:base}. This thickening is necessary when the dimensionality of \dbase\ is less than the dimensionality of the target display. For example, a \dbasepoint\ that is 0D in \dbase\ cannot be represented on screen unless it is thickened to 2D to encode the connectivity of the points in \gfiber\ that visually represent the record at \dbasepoint. The \vindex\ mapping is critical to interactive visualizations as it is the map from a region on screen to the data associated with that region. One example is to fill in details in a hover tooltip, another is to convert region selection on \gbase\ to a query on the data to access the corresponding record components on \dbase.

@@

\begin{figure}[htb]
  \centering
  \includegraphics[width=1\columnwidth]{monoid_equivariant.png}
  \caption{Equivariant \vchannel\ maps the weather data to corresponding symbols such that the relative order of the symbols is the same as the data as shown in the top two panels. In the bottom right panel the ordering of \textit{rain} and \textit{storm} is swapped in the symbols, meaning that the \vchannel\ is invalid; in the bottom left panel, \textit{storm} is instead mapped to the same symbol as \textit{rain} so that the ordering is preserved.}
  \label{fig:math:artist:nu}
\end{figure}

\begin{table}[H]
  \renewcommand{\arraystretch}{2}
  \begin{tabulary}{\columnwidth}{|llL|}\hline
      \textbf{scale} & \textbf{group} & \textbf{constraint}\\ \hline
      nominal & permutation &  $\text{if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)$\\
      ordinal &  monotonic & $\text{if } \delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)$\\
      interval &  translation &  $\vchannel (x + c) = \vchannel(x) + c$ \\
      ratio &  scaling &  $\vchannel(xc) = \vchannel(x)*c $\\ \hline
  \end{tabulary}
  \label{tab:math:artist:nu}
\end{table}
The equivariance constraints for the Steven measurement scales\cite{stevensTheoryScalesMeasurement1946} are derived from the mathematical group structure of the scales. For monoid actions, the equivariance constraint is that relative order is preserved, as shown in the top two panels of \autoref{fig:math:artist:nu}.

>>>>

%% the only exception to this rule is the \firstsection command
\firstsection{Introduction}
\maketitle
%% this is still a mess, come back to it
Models for describing visualizations and visualization tools are deeply tied to to the domains in which the visualizations are made \cite{friendlyBriefHistoryData2006a,pousmanCasualInformation2007}, and these tools are often made from building block libraries \cite{wongsuphasawatNavigatingWideWorld2021} such as the Python visualization library Matplotlib. Because building block libraries often support so many domains, the API can grow organically in ways that are incoherant and brittle. To adapt to modern domain library development needs, this paper presents a functional model of the constraints building block software needs to satisfy so that a visualization, however it is specified in the library built out of these blocks, is guaranteed to be a faithful representation of the data. Our model, which we call the  topological equivariant artist model (TEAM), provides a means to express the constraints of preserving the data continuity in the graphic and faithfully translating the properties of the data variables into visual variables.

\begin{figure}
  \includegraphics[width=\columnwidth]{figures/continuity.png}
  \caption{Continuity describes how the records in the dataset, in this case the rows in the table, are connected to each other. The line plot is an invalid visual representation of the table because it implies that the records are connected to each other in some way. On the other hand the bar plot is valid because it preserves the records as independent data points.}
  \label{fig:intro:continuity}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figures/equivariance.png}
  \caption{Equivariance means that transformations on the data side and on the visual side are equal. In this example, the temperature is doubled via the arrows representing the doubling action, and the plot of the temperature is doubled via the arrows on the visual side.}
  \label{fig:intro:equivariance}
\end{figure}

Motivated by the challenge of rearchitecturing a general purpose visualization library, we developed the Topological Equivariant Artist Model (TEAM) to express the constraints of preserving data continuity and translating data properties into a visual representation. The contribution of TEAM is 
\begin{enumerate}
  \item topology preserving relationship between data and graphic via continuous maps
  \item property preservation from data component to visual representation as monoid action equivariant maps
  \item functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model.
\end{enumerate}

\section{Related Work}
%
% My best paraphrasing of the main idea would be that this paper takes the ideas of going from data to graphic along with some important constraints that people generally agree are important and formalizes them into a topological model, then shows how that model can map into a concrete implementation. The important part of this model is that it models the data sufficiently abstractly that it's able to work with basically any form or type of data. The sentence starting at line 489 seems to be the big "aha" to me, basically saying something like "if you implement things such that the code equivalents of these terms are separated in some specific way, then you get these important formally-describable properties". Is that right? If so, I'm left very curious about what alternative (possibly already-existing?) Artist implementations there might be that seem to work, but don't decompose things in this way, and therefore don't have these important properties. Or maybe the existing implementations do have these properties but aren't general enough? I know there are space constraints etc but if there was some way of showing how the current Matplotlib artists work and how they fall short of this model, that would be very enlightening I think.

\section{Topological Equivariant Artist Model}
\label{sec:math}
We introduce the notion of an artist $\mathscr{\vartist}$ as an equivariant map
\begin{equation}
    \label{eq:artist}
    \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
\end{equation}
from data $\mathscr{\dtotal}$ (\autoref{sec:math:data}) to graphic $\mathscr{\gtotal}$ (\autoref{sec:math:graphic}) fiber bundles. We decompose the artist into an indexing map from graphic to data (\autoref{sec:math:graphic:base}), a map from data components to visual components (\autoref{sec:math:artist:nu}), and map from visual components to graphic (\autoref{sec:math:artist:q}). In this section we discuss the formal properties of these fiber bundles and maps such that they can be used to specify an implementation, for example the one prototyped in \autoref{sec:code}.

@@

\subsubsection{Fiber Space: Variables}
\label{sec:math:data:fiber}
To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES,spivakDatabasesAreCategories2010} that binds the components of the fiber to variable names. Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_m\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset. He then defines the single variable set \fttype, which is \ftotal\ restricted to objects of type \ftype\ bound to variable name \fname. The \fttype\ lookup is by name to specify that every component is distinct, since multiple components can have the same type \ftype. Given \fsection, the fiber for a one variable dataset is
\begin{equation}
    \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
\end{equation}
where \fsection\ is the schema binding variable name \fname\ to its datatype \ftype. A dataset with multiple variables has a fiber that is the cartesian cross product of $\ftotal_{\fsection}$ applied to all the columns:
\begin{equation}
F = \ftotal_{\fsection(\fname_{1})}\times \ldots \ftotal_{\fsection(\fname_{i})} \ldots\times \ftotal_{\fsection(\fname_{n})}
\end{equation}
which is equivalent to 
\begin{equation}
    \label{eq:math:data:fiber:decompose}
    \dfiber= \dfiber_{0} \times \ldots \times \dfiber_{i}\times\ldots\times \dfiber_{n}
\end{equation}
which allows us to decouple \dfiber\ into components $\dfiber_i$. Each component of \dfiber\ is a dimension of the topological fiber space and is specified by a tuple of the form (\fname, \ftype, \(\ftotal_{\fsection(\fname)}\)). 
\begin{figure}[htb]
  \centering 
  \includegraphics[width=\columnwidth]{fiber.png}
  \caption{These two datasets have the same base space \dbase. The plane is a representation of the fiber $\dfiber=\reals\times\reals$ for the variables (time, temperature) from \autoref{fig:intro:equivariance}, while the cube is the fiber $\reals\times\realsp\times\reals$ associated with (time, wind=(speed, direction))}
  \label{fig:math:data:fiber}
\end{figure}

In \autoref{fig:math:data:fiber} the plane fiber has components (\emph{time}, \texttt{datetime}, \reals) and (\emph{temperature}, \texttt{float}, \reals), while the cube fiber has components (\emph{time}, \texttt{datetime}, \reals) and (\emph{wind}, \texttt{wind}, \(\realsp\times\reals)\)) which encodes (\emph{speed}, \emph{direction}).

@@

A monoid \cite{Monoid2021} $\monoid$ is a set with an associative binary operator $\ast:\monoid \times \monoid\rightarrow \monoid$. A monoid has an identity element $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$. As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,nlab:action} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties of associativity and identity. As with the fiber \dfiber\, the total monoid space \monoid\ is the cartesian product
\begin{equation}\label{eq:math:data:monoid:decompose}
\monoid= \monoid_{0} \times \ldots \times \monoid_{i}\times \ldots \times\ldots \monoid_{n}
\end{equation}
of each monoid $\monoid_{i}$ on $\dfiber_{i}$.  The monoid is also added to the specification of the fiber $(\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)$

Defining the monoid actions on the components serves as the basis for identifying the invariance\cite{kindlmannAlgebraicProcessVisualization2014} that must be preserved in the visual representation of the component. Monoids are commonly found in functional programming because a property of monoids is that the components can be composed into complex transformation pipelines \cite{yorgeyMonoidsThemeVariations}. For example, a component that converts \(\dfiber_i=time\) with \(\monoid_i\) to hours can be combined with a component that shifts \(time\) by 2 hours to yield \(hours_2\) with the same \(monoid_i\).

\subsubsection{Base Space: Continuity}
\label{sec:math:data:base}

 The base space \dbase\ acts as an indexing space, as emphasized by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, to express how the records in \dtotal\ are connected to each other. This is similar the notion of structural \textit{keys} with associated \textit{values} proposed by Munzner\cite{munznerVisualizationAnalysisDesign2014}, but our model treats keys as a pure reference to topology. Decoupling the keys from their semantics allows the metadata to be altered facilitates encoding of data where the independent variable may not be clear; for example the growth of a plant is dependent on both time and sunlight, and changing the changing the coordinate system or time resolution should have no effect on how the records are connected to each other.

  \begin{figure}[htb]
    \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
    \includegraphics[width=\columnwidth]{base.png}
    \caption{These two datasets have the same fiber space $\dfiber=\reals\times\reals$. For example, the left dataset is a set of discrete (\textit{time}, \textit{temperature}) records while the right is a 1D continuous function over the interval $\left[0,1\right]$ with return values of the form (\textit{time}, \textit{temperature}).}
    \label{fig:math:data:base}
   \end{figure}
 As illustrated in ~\autoref{fig:math:data:base}, \dbase\ can have any number of dimensions, can be continuous or discrete, and is somewhat independent of the dimensions of the fiber. Every $\dbasepoint \in \dbase$ has a corresponding fiber $\dfiber_k$ because \dbase\ is the quotient space\cite{QuotientSpaceTopology2020,aurouxMath131Introduction} of \dtotal. As with \autoref{eq:math:data:fiber:decompose} and \autoref{eq:math:data:monoid:decompose}, we can decompose the total space into component bundles
 \begin{equation}\label{eq:math:data:base:decompose}
     \pi:\dtotal_1\oplus\ldots\oplus \dtotal_i \oplus\ldots \oplus \dtotal_n \rightarrow \dbase
 \end{equation}
 such that \(\monoid_{i}\) acts on component bundle \(\dtotal_i\). The \dbase\ remains the same because the connectivity of records does not change just because there are fewer components in each record. By encoding this continuity in the model as \dbase\, the data model now explicitly carries information about its structure such that the implicit assumptions of the visualization algorithms are now explicit. The explicit topology is a concise way of distinguishing visualizations that appear identical, for example heatmaps and images.

 \subsubsection{Section: Values}
 \label{sec:math:data:section}

 \begin{figure}[htb]
  \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
  \includegraphics[width=\columnwidth]{fiberbundle.png}
  \caption{Each section in the fiber bundle is a unique continuous map from base space to fiber encoding a set of records in the dataset. For example, the two sections $\dsection^{(1)}$ and $\dsection^{(2)}$ each encode a timeseries from a different weather station.}
  \label{fig:math:data:section}
 \end{figure}

 While the projection function $\pi:\dtotal \rightarrow\dbase$ ties together the base space \dbase\ with the fiber \dfiber, a section $\dsection: \dbase\rightarrow \dtotal$ encodes a dataset. A section function takes as input location $\dbasepoint \in \dbase$ and returns a record $\delement \in \dtotal$. For any fiber bundle, there exists a map
 \begin{equation}
     \begin{tikzcd}
         \dfiber \arrow[r, hook] & \dtotal \arrow[d, "\pi"'] \\
                           & \dbase \arrow[u, "\dsection"', bend right]
     \end{tikzcd}
 \end{equation}
  such that $\pi(\dsection(\dbasepoint)) = \dbasepoint$. The set of all global sections is denoted as $\Gamma(\dtotal)$. As illustrated in \autoref{fig:math:data:section}, the section is a continuous mapping from a location $\dbasepoint \in \dbase$ on the base space to a record $\delement \in \dfiber$ in the fiber. Assuming a trivial fiber bundle $\dtotal = \dbase \times \dfiber$, a section
 \begin{equation}
     \label{eq:section_return}
     \dsection(\dbasepoint) = (\dbasepoint, (g_{\dfiber_{0}}(\dbasepoint), \ldots, g_{\dfiber_{n}}(\dbasepoint)))
 \end{equation}
 returns a record for each \dbasepoint. The index function $g: \dbase \rightarrow \dfiber$ into each fiber component returns a value in the component. This formulation of the section also holds on locally trivial sections of a non-trivial fiber bundle. As with \autoref{eq:math:data:fiber:decompose} and \autoref{eq:math:data:base:decompose}, \dsection\ can be decomposed into components 
 \begin{equation}
 \label{eq:math:data:section:decompose}
 \dsection= (\dsection_0,\ldots, \dsection_i, \dots, \dsection_n) 
 \end{equation}
 where each section $\dsection_i$ maps into a record on a component $\dfiber_i \in \dfiber$. This allows for accessing the data component wise in addition to accessing the data in terms of its location over \dbase.

 \subsubsection{Sheafs}
 \label{sec:math:data:sheaf}
A sheaf is a mathematical structure for defining collections of objects\cite{ghristElementaryAppliedTopology2014, ghristHomologicalAlgebraData2018,urbanikBriefIntroductionSchemes} on mathematical spaces. On the fiber bundle \dtotal, we can describe a sheaf as the collection of local sections $\iota^{*}\dsection$
 \begin{equation}
     \label{eq:sheaf}
     \begin{tikzcd}[column sep=huge]
         \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             & \dtotal \arrow[d, "\pi"']                  \\
         U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] & \dbase \arrow[u, "\dsection"', bend right]
     \end{tikzcd}
 \end{equation}
which are sections of \dtotal\ pulled back over local neighborhood $U \subset \dtotal$ via the inclusion map \(\iota:\dtotal\rightarrow U\). The collation of sections enabled by sheafs is necessary for navigation techniques such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006} and dynamically updated visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995}

@@

Just as the \dbase\ encodes the connectivity of the records in the data, we propose an equivalent \gbase\ that encodes the connectivity of the rendered elements of the graphic. Formally, we require that \dbase\ be a deformation retract\cite{RetractionTopology2020} of \gbase\ so that \dbase\ and \gbase\ have the same homotopy. The surjective map $\vindex: \gbase \rightarrow \dbase$ 
\begin{equation}
    \begin{tikzcd}
        \dtotal \arrow[d, "\pi"'] & \gtotal \arrow[d, "\pi"'] \\
        \dbase                   & \gbase \arrow[l, "\vindex"']
    \end{tikzcd}
\end{equation}
goes from region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\gbasepoint$. While \gbase\ must have the same continuity as \dbase\, it is sometimes the  thickened version shown in \autoref{fig:math:graphic:base}. This thickening is necessary when the dimensionality of \dbase\ is less than the dimensionality of the target display. For example, a \dbasepoint\ that is a point in 0D \dbase\ cannot be represented on screen unless it is thickened to 2D to encode the connectivity of the points in \gfiber\ that visually represent the record at \dbasepoint. The \vindex\ mapping is critical to interactive visualizations as it is the map from a region on screen to the data associated with that region. One example is to fill in details in a hover tooltip, another is to convert region selection on \gbase\ to a query on the data to access the corresponding record components on \dbase.

@@

\begin{figure}[htb]
  \centering
  \includegraphics[width=1\columnwidth]{figures/partial_fixed.png}
  \caption{The equivariance constraint is not met in the first figure because \(rain\geq storm\), but \(\vchannel(rain) \leq \vchannel(storm)\) as shown by the crossed arrows. This is fixed by mapping \textit{storm} to the same element as \textit{rain}, such that \(\vchannel(rain) \geq \vchannel(storm)\).}
  \label{fig:math:artist:nu}
\end{figure}

\begin{table}[H]
  \renewcommand{\arraystretch}{2}
  \begin{tabulary}{\columnwidth}{|llL|}\hline
      \textbf{scale} & \textbf{group} & \textbf{constraint}\\ \hline
      nominal & permutation &  $\text{if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)$\\
      ordinal &  monotonic & $\text{if } \delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)$\\
      interval &  translation &  $\vchannel (x + c) = \vchannel(x) + c$ \\
      ratio &  scaling &  $\vchannel(xc) = \vchannel(x)*c $\\ \hline
  \end{tabulary}
  \label{tab:math:artist:nu}
\end{table}
We can state the conditions on \vchannel\ such that they cover the Stevens measurement scales\cite{stevensTheoryScalesMeasurement1946}, as defined in \autoref{tab:math:artist:nu}. This is because Stevens’ defined the measurement scales in terms of their mathematical group structure and a group is a monoid with inverses\cite{remlingAlgebraMath5353}. An example of monoid action equivariance is the preservation of partial ordering illustrated in \autoref{fig:math:artist:nu}.
